AFQ-Insight

### 2019-11-16 reports/manuscript/abstract.tex
Add methods section to manuscript

\begin{abstract}

\adam{This is a copy of the abstract submitted for SfN 2018}

The white matter contains long-range connections between different brain regions. Tractometry uses diffusion-weighted magnetic resonance imaging (dMRI) data to quantify tissue properties along the trajectories of these connections in the human brain in vivo\cite{yeatman2012tract}. In previous work, the results of tractometry were usually analyzed using mass univariate approaches: group comparisons or regression models computed separately for each point along every one of the tracts.  Alternatively, tissue properties such as fractional anisotropy (FA) and mean diffusivity (MD) were computed for a specific tract of interest based on an a priori hypothesis. In the present work, we developed a method based on the sparse group lasso\cite{simon2013sparse} that takes into account all of the tissue properties measured in all of the tracts, and selects informative features by enforcing sparsity, both at the level of individual tracts and tissue properties, but also across the entire set of tracts and all of the measured tissue properties. The sparsity penalties for each of these constraints is identified using a nested cross-validation scheme. Using data from a previous study that measured dMRI in patients with amyotrophic lateral sclerosis (ALS) and matched controls\cite{sarica2017corticospinal}, we demonstrate that this method is accurate, exceeding the previous results, that were based on a priori feature selection, in classifying patients and controls, with ~84\% accuracy, an area under the receiver operating characteristic curve of 0.93, and an average precision of 0.95. Moreover, our method automatically identifies as important for this classification the parts of the white matter known to be affected by ALS within the corticospinal tract. In a regression setting, data from another previous study\cite{yeatman2014lifespan} can be used to accurately predict ``brain age''. Thus, this multivariate analysis approach both (a) achieves high cross-validated accuracy for precision medicine applications of dMRI data and (b) identifies relevant features of brain anatomy to further our neuroscientific understanding of clinical disorders.

\end{abstract}

>>>>

\section*{Abstract}

The white matter contains long-range connections between different brain
regions. Tractometry uses diffusion-weighted magnetic resonance imaging (dMRI)
data to quantify tissue properties along the trajectories of these connections
in the human brain in vivo\cite{yeatman2012tract}. In previous work, the results
of tractometry were usually analyzed using mass univariate approaches: group
comparisons or regression models computed separately for each point along every
one of the tracts.  Alternatively, tissue properties such as fractional
anisotropy (FA) and mean diffusivity (MD) were computed for a specific tract of
interest based on an a priori hypothesis. In the present work, we developed a
method based on the sparse group lasso\cite{simon2013sparse} that takes into
account tissue properties measured in all of the tracts, and selects informative
features by enforcing sparsity, both at the level of individual tracts and
tissue properties, but also across the entire set of tracts and all of the
measured tissue properties. The sparsity penalties for each of these constraints
is identified using a nested cross-validation scheme. Using data from a previous
study that measured dMRI in patients with amyotrophic lateral sclerosis (ALS)
and matched controls\cite{sarica2017corticospinal}, we demonstrate that this
method is accurate, exceeding the previous results, that were based on a priori
feature selection, in classifying patients and controls, with ~84\% accuracy, an
area under the receiver operating characteristic curve of 0.93, and an average
precision of 0.95. Moreover, our method automatically identifies as important
for this classification the parts of the white matter known to be affected by
ALS within the corticospinal tract. In a regression setting, data from another
previous study\cite{yeatman2014lifespan} can be used to accurately predict
``brain age''. Thus, this multivariate analysis approach both (a) achieves high
cross-validated accuracy for precision medicine applications of dMRI data and
(b) identifies relevant features of brain anatomy to further our neuroscientific
understanding of clinical disorders.

### 2019-11-16 reports/manuscript/paper.tex
First pass of several sections of the Introduction.
Haven't quite gotten to the point where I describe what/why we're doing here.

But it's a start.

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{AFQ-Insight: multidimenstional analysis and detection of informative features in diffusion MRI measurements of human white matter} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Adam Richie-Halford\textsuperscript{1,*},
Jason Yeatman\textsuperscript{2},
Noah Simon\textsuperscript{3},
Ariel Rokem\textsuperscript{4},
\\
\bigskip
\textbf{1} Department of Physics, University of Washington, Seattle, WA, 98105, USA
\\
\textbf{2} Department of Speech and Hearing Sciences, University of Washington, Seattle, WA, 98105, USA
\\
\textbf{3} Department of Biostatistics, University of Washington, Seattle, WA, 98105, USA
\\
\textbf{4} e{S}cience Institute, University of Washington, Seattle, WA, 98105, USA
\\
\bigskip

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Multidimensional analysis and detection of informative features in diffusion MRI measurements of human white matter} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Adam Richie-Halford\textsuperscript{1,*},
Jason Yeatman\textsuperscript{2},
Noah Simon\textsuperscript{3},
Ariel Rokem\textsuperscript{4},
\\
\bigskip
\textbf{1} Department of Physics, University of Washington, Seattle, WA, 98105, USA
\\
\textbf{2} Graduate School of Education and Division of Developmental and Behavioral Pediatrics, Stanford University, Stanford, CA, 94305, USA
\\
\textbf{3} Department of Biostatistics, University of Washington, Seattle, WA, 98105, USA
\\
\textbf{4} e{S}cience Institute, University of Washington, Seattle, WA, 98105, USA
\\
\bigskip

### 2019-11-16 reports/manuscript/sections/conclusion.tex
Add methods section to manuscript

\section{Conclusion}

>>>>

\section*{Conclusion}

We present here a novel method for analysis of dMRI tractometry data. This
method relies on the Sparse Group Lasso \cite{simon2013sparse} to provide both
acurate prediction of the phenotypic properties of individual subjects based on
their dMRI data, but also provides interpretable results by identifying the
features that are important across subjects to make these predictions. The
method is broadly applicable: it performs well in predicting both continuous
variables, such as biological age, as well as discrete variable, such as whether
a person is a patient or a healthy control. In both of these cases, SGL
out-performs previous algorithms that have been developed for these tasks. In
addition, the nested cross-validation approach that we have developed tunes the
degree of sparseness required by the algorithm, such that both very local
phenomena, as well as widely distributed phenomena are accurately captured.

Specifically, we demonstrated that the algorithm correctly detects the fact that
 ALS, which is a disease of lower motor neurons, is localized to the
 cortico-spinal tract. This recapitulates the results of previous analysis of
 these same data, using a targetted ROI-based approach. In contrast, in analysis
 of biological age, the coefficients identified by the algorithm are very widely
 distributed.

Taken together, these results demonstrate the promise of a group regularized
regression approach. Even at the scale of dozens of subjects, the results
provided by SGL are both accurate and interpretable. We expect that additional
data will improve the performance of these algorithms. Neuroscience has entered
an era in which large consortium efforts are putting together large datasets
that can be analyzed using this approach. Future work will apply this method to
these datasets. For example, to the Human Connectome Project dataset
\cite{VanEssen2012}

These results also motivate extensions of the method using more sophisticated
cost functions. For example, the fused sparse group lasso (FSGL)
\cite{zhou2012} extends SGL to enforce additional spatial structure: smoothness
in the variation of diffusion metrics along the tracts. As brain measurements
include additional structure (for example, bilateral symmetry), future work
could also incorporate overlapping group membership for each entry in the tract
profiles. For example, a measurement could come from the corpus callosum, but
also from the right hemipshere. This would also require extending the cost
function used here to incorporate these constraints.

The method is packaged as open-source software called AFQ-Insight that is openly
available. The sofware integrates within a broader automated fiber
quantification software ecosystem: AFQ \cite{yeatman2012tract}, which extracts
tractometry data from raw dMRI, as well as AFQ-Browser, which visualizes
tractometry data and facilitates sharing of the results of dMRI studies
\cite{yeatman2018browser}.

### 2019-11-16 reports/manuscript/sections/introduction.tex

Diffusion-weighted Magnetic Resonance Imaging (dMRI) provides a unique view into
the physical properties of the connections that comprise the brain white matter
\cite{wandell2016clarifying}. Among the benefits of the method are the fact that
it is non-invasive, and feasible to conduct in clinical scanners, and in wide
variety of populations: healthy adults, but also clinical populations, as well
as young or elderly subjects. While the measurements are usually conducted with
voxels at the millimeter scale, within the diffusion time usually used, water
molecules within each voxel can probe the structure of white matter at the
micrometer scale, providing important information about the physical structure
of the white matter, such as the distribution of fiber orientations within each
voxel \cite{Frank2002-iz, Frank2001-xf}. With more sophisticated measurements
that include several diffusion-weightings, more information can be extracted
about the composition of different tissue type in the voxel
\cite{Jeurissen2014-ab, Kaden2016-ho}. In addition, phenomenological
descriptions of the diffusion properties, such as its mean diffusivity or
fractional anisotropy can be derived from dMRI measurements
\cite{Pierpaoli1996-vj}. Some of these descriptions are ambiguous in terms of
their interpretation, as variance in these properties can relate to several
different underlying physical causes. For example, fractional anisotropy can
decrease both because of demyelination, but also because of decreased density,
or because of partial volume with other tissue types or between crossing fiber
populations \cite{Beaulieu1996-fn, Beaulieu2002-tl}. Even so, these descriptions
have proven useful in characterizing individual differences or differences
between populations \ariel{citation needed}.

@@

In the present work, we present a novel framework for analysis of tractometry
that simultaneously selects the features for analysis, and fits a model to these
features. Rather than a comparative approach, this framework takes a predictive
approach to inference \cite{Breiman2001}. We use a linear modeling approach,
which aims to predict phenotypical variance in a group of subjects, based on a
linear combination of the features estimated with tractometry. Using this
approach, we need to first deal with the large and asymmetric dimensionality of
the data: tractometry data usually has many more features (i.e., number of
measurements per individual) than samples (number of subjects), which makes
inferences from the data about phenotypical differences between individuals
ill-posed. This regime is the target of several statistical learning techniques,
and is often solved by various forms of regularization. For example, Tikhonov
regularization shrinks the solution such that the sum of squared contributions
from the individual features are minimized \cite{Hoerl2000-ij}. Another solution
to the problem is provided by the Lasso algorithm, which instead minimizes the
sum of the absolute values of contributions of each feature
\cite{Tibshirani1996-qs}. This tends to shrink to zero the contributions of many
of the features, providing results that are both accurate and interpretable.
When additional structure is available in the organization of the data,
regularization algorithms can take advantage of this structure. For example, if
the features lend themselves to a natural division into different groups, the
group lasso (GL) can be used to select groups of features, rather than
individual features \cite{Yuan2006-ky}. The Sparse Group Lasso (SGL) elaborates
on this idea by providing control both of group sparsity, as well as overall
sparsity of the solutions \cite{simon2013sgl}. Because the features measured
with tractomery lend themselves to grouping based on the tracts from which each
measurement is taken, GL and SGL provide a useful tool linear model fitting in
problems of this form. Using linear regression as the basis for analysis is
useful for both regression tasks, in which a continuous measure (e.g., age) is
modeled, as well as for classification tasks, where a discrete property is
modeled (e.g., ). Moreover, use of SGL in particular helps derive informative
properties of the human white matter, both in cases where the underlying causes
of variance are sparsely concentrated (e.g., in one particular white matter
tract) and in cases where biological substrates of differences between
individuals are more widespread. This is because SGL is controlled by
meta-parameters that set the degree to  which and that can be set empirically,
based on the data, in a process of nested cross-validation. We demonstrate these
properties of the algorithm in data from previously published studies
\cite{sarica2017corticospinal, yeatman2014lifespan}.

>>>>

Diffusion-weighted Magnetic Resonance Imaging (dMRI) provides a unique view into
the physical properties of the connections that comprise the brain white matter
\cite{wandell2016clarifying}. Among the benefits of the method are the fact that
it is non-invasive, and feasible to conduct in clinical scanners, and in wide
variety of populations: healthy adults, but also clinical populations, as well
as young or elderly subjects. While the measurements are usually conducted with
voxels at the millimeter scale, within the diffusion time usually used, water
molecules within each voxel can probe the structure of white matter at the
micrometer scale, providing important information about the physical structure
of the white matter, such as the distribution of fiber orientations within each
voxel \cite{Frank2002-iz, Frank2001-xf}. With more sophisticated measurements
that include several diffusion-weightings, more information can be extracted
about the composition of different tissue type in the voxel
\cite{Jeurissen2014-ab, Kaden2016-ho}. In addition, phenomenological
descriptions of the diffusion properties, such as its mean diffusivity or
fractional anisotropy can be derived from dMRI measurements
\cite{Pierpaoli1996-vj}. Some of these descriptions are ambiguous in terms of
their interpretation, as variance in these properties can relate to several
different underlying physical causes. For example, fractional anisotropy can
decrease both because of demyelination, but also because of decreased density,
or because of partial volume with other tissue types or between crossing fiber
populations \cite{Beaulieu1996-fn, Beaulieu2002-tl}. Even so, these descriptions
have proven useful in characterizing individual differences or differences
between populations \cite{}.

@@

In the present work, we present a novel framework for analysis of tractometry
that simultaneously selects the features for analysis, and fits a model to these
features. Rather than a comparative approach, this framework takes a predictive
approach to analysis \cite{Yarkoni2017-bi}. We use a linear modeling approach,
which aims to predict phenotypical variance in a group of subjects, based on a
linear combination of the features estimated with tractometry. Using this
approach, we need to first deal with the large and asymmetric dimensionality of
the data: tractometry data usually has many more features (i.e., number of
measurements per individual) than samples (number of subjects), which makes
inferences from the data about phenotypical differences between individuals
ill-posed. This regime is the target of several statistical learning techniques,
and is often solved by various forms of regularization. For example, Tikhonov
regularization shrinks the solution such that the sum of squared contributions
from the individual features are minimized \cite{Hoerl2000-ij}. Another solution
to the problem is provided by the Lasso algorithm, which instead minimizes the
sum of the absolute values of contributions of each feature
\cite{Tibshirani1996-qs}. This tends to shrink to zero the contributions of many
of the features, providing results that are both accurate and interpretable.
When additional structure is available in the organization of the data,
regularization algorithms can take advantage of this structure. For example, if
the features lend themselves to a natural division into different groups, the
group lasso (GL) can be used to select groups of features, rather than
individual features \cite{Yuan2006-ky}. The Sparse Group Lasso (SGL) elaborates
on this idea by providing control both of group sparsity, as well as overall
sparsity of the solutions \cite{simon2013sgl}. Because the features measured
with tractomery lend themselves to grouping based on the tracts from which each
measurement is taken, GL and SGL provide a useful tool linear model fitting in
problems of this form. Using linear regression as the basis for analysis is
useful for both regression tasks, in which a continuous measure (e.g., age) is
modeled, as well as for classification tasks, where a discrete property is
modeled (e.g., ). Moreover, use of SGL in particular helps derive informative
properties of the human white matter, both in cases where the underlying causes
of variance are sparsely concentrated (e.g., in one particular white matter
tract) and in cases where biological substrates of differences between
individuals are more widespread. This is because SGL is controlled by
meta-parameters that set the degree to  which and that can be set empirically,
based on the data, in a process of nested cross-validation. We demonstrate these
properties of the algorithm in data from previously published studies
\cite{sarica2017corticospinal, yeatman2014lifespan}.

### 2019-11-16 reports/manuscript/sections/results.tex

\subsection*{Regression for age prediction}
The WH dataset used for continuous regression contains data from 76 healthy
subjects, ranging between 6 years and 50 years of age
\cite{yeatman2014lifespan}. SGL fit to tractometry-extracted features in these
subjects accurately predicts the age of the subjects. The model reaches a mean
absolute error of 3.6 years (Figure \ref{fig:regress-results}), which is lower
than the results of a recent study that predicted age in a large sample, based
on TBSS-extracted diffusion features \cite{Richard2018-ux}. The model weights
in this case are distributed over many different tracts and dMRI tissue
properties (Figure \ref{}) This demonstrates that SGL is not coerced to produce overly sparse
results when a more accurate model requires a dense selection of features.

>>>>

\subsection*{Regression for age prediction}

The WH dataset used for continuous regression contains data from 76 healthy
subjects, ranging between 6 years and 50 years of age
\cite{yeatman2014lifespan}. SGL fit to tractometry-extracted features in these
subjects accurately predicts the age of the subjects. The model reaches a mean
absolute error of 3.6 years (Figure \ref{fig:regress-results}), which is lower
than the results of a recent study that predicted age in a large sample, based
on TBSS-extracted diffusion features \cite{Richard2018-ux}. The model weights in
this case are distributed over many different tracts and dMRI tissue properties
(Figure \ref{fig:regress-beta}) This demonstrates that SGL is not coerced to
produce overly sparse results when a more accurate model requires a dense
selection of features.
